Optimizing Structural Properties Of Neural Networks With Genetic Algorithms

### 2015-08-06 paper/paper.tex
some text cuts

We chose the same rate based network architecture
that is used in the original FORCE publication \citep{FORCE} (therein denoted as
architecture A) and also in \citep{RM}, where our notation mainly stems
from. So we look at a network $\mathcal{N}$ with $N$ internal neurons and
states $x_i$ for $1 \le i \le N$ that loosely represent the neurons'
membrane potential. The firing rate $r_i$ of the $i$-th internal neuron is
given by $r_i = \tanh x_i$.  Furthermore, the network contains $R$ readout
neurons with states $z_i$ for $1 \le i \le R$, where
\begin{equation} 
    z_i = \sum_{j=1}^{N}\omega\Ix{read}_{ij}\,r_j
\end{equation}
with the readout weights $\omega\Ix{read}$.
Taking into account both internal dynamics as well as an external feedback
pathway, the dynamics $t \mapsto x_i(t)$ of the single neurons are governed
by \footnote{~Note that we did not include external input signals in this
formula, as is done in \citep{FORCE} and \citep{RM}. We in fact enabled
inputs in our code base, but did not use them for the main results.}
\begin{equation}
    \dot{x}_i(t) = -x_i(t) + \sum_{i=1}^N \omega\Ix{rec}_{ij}\, r_j(t) + 
                   \sum_{i=1}^R \omega\Ix{fb}_{ij}\, z_j(t)~,
\end{equation}
where we introduce the internal recurrent synapse weights
$\omega\Ix{rec}_{ij}$ and the feedback synapse weights
$\omega\Ix{fb}_{ij}$. 
We solved this system of differential equations by
a simple Newton method with integration time steps $dt = 0.01$, which was
also used in \citep{FORCE}. 
%(How we constructed $\omega\Ix{rec}$ and
%$\omega\Ix{fb}$ will be described in \emph{Parametric Random Networks})

\paragraph{Learning} Learning took place by stepwisely adapting the weights
$\omega\Ix{read}$ so that the dynamics $t\mapsto r_i(t)$ of the readout
neurons should eventually match a given periodic target pattern $t\mapsto
r'_i(t)$, henceforth called a \emph{task}, as accurately as possible. To
update $\omega\Ix{read}$ we chose a variant of the FORCE algorithm
\citep{FORCE} that rapidly modifies the feedback loop to keep the errors
$|r_i(t) - r'_i(t)|$ small from the beginning on.

While the authors of \citep{FORCE} update the weights $\lambda\Ix{read}$ in
fixed intervals (every second time step), we used an adaptive mechanism to
determine how often learning should occur. This was initially introduced
for performance reasons, as the single learning steps proved to be the main
computational bottleneck, but we also found that the long time network
dynamics tended to be slightly more accurate for adaptive learning steps.

\paragraph{Network Genotypes} Since we are not
interested in single network properties but rather want to find out which
structural elements of networks are important for learning success, we
introduced  \emph{network genotypes}.
Such a genotype $G$ consists of a set $\Theta$ of parameters that determine
which concrete networks are created when $G$ is expressed in a single
phenotype. Technically $G$ may be understood as a stochastic network
generator that returns a network $\mathcal{N}=G(\lambda)$ for every seed
$\lambda \in \mathbb{N}$.
The genotypes' parameters $\theta\in\Theta$ were either integer or floating
point values and can be classified as affecting either (1) the topology of
$\omega\Ix{rec}$ or (2) the weight values of $\omega\Ix{rec}$ and
$\omega\Ix{fb}$. Important examples for (1) are the networks size $N$, the
occupation probability $p$ (for random Erdös-Renyi networks), and the
neighborhood range $k$ (for ring topologies). We furthermore introduced ratio parameters
that allowed us to interpolate between different topologies. Optimized
parameters of type (2) were the feedback strength $\feedback$ (with
$\omega\Ix{fb}\propto \feedback$) and the gain value $\gain$ (with
$\omega\Ix{rec} \propto \gain$).

\paragraph{Fitness} The next step was to define the \emph{fitness} of a given
genotype when confronted with a diverse range of tasks, e.g. waves of
different frequencies. To express this formally, we introduced
\emph{challenges}: For a given seed value $\lambda$, the challenge $C$
yields a task $r' = C(\lambda)$ according to some (parametrized)
distribution. If we now make use of a success function $S(\mathcal{N},
r')\in\{0, 1\}$ that decides whether the network $\mathcal{N}$
has successfully reproduced the desired output $r'$, the fitness $F_C(G)$
of a genotype $G$ for a challenge $C$ may reasonably be defined as the
expectation value of $\lambda \mapsto S(G(\lambda), C(\lambda))$. For fixed
challenges, the main task of optimizing the genotype is now reduced to
finding parameters $\Theta$ that maximize $F_C$.

@@

\paragraph{Choice of Parameters and Algorithmic Quantities}
For all of the simulations we chose the size of the networks to be fixed
with $N=100$. Though increasing $N$ generally improved the quality of the
learning results, the computational cost increased proportional to $N^2$,
which made using networks as large as in the original FORCE publication
(between $500$ and $20000$) utterly unusable for our purposes.

The mutation rate of the parameters $\Theta$ of the genotypes\dots

>>>>

We chose the same rate based network architecture
that is used in the original FORCE publication \citep{FORCE} (therein denoted as
architecture A) and also in \citep{RM}, where our notation mainly stems
from. So we look at a network $\mathcal{N}$ with $N$ internal neurons and
states $x_i$ for $1 \le i \le N$ that loosely represent the neurons'
membrane potential. The firing rate $r_i$ of the $i$-th internal neuron is
given by $r_i = \tanh x_i$.  Furthermore, the network contains $R$ readout
neurons with states $z_i$ for $1 \le i \le R$, where
\begin{equation} 
    z_i = \sum_{j=1}^{N}\omega\Ix{read}_{ij}\,r_j
\end{equation}
with the readout weights $\omega\Ix{read}$.
Taking into account both internal dynamics as well as an external feedback
pathway, the dynamics $t \mapsto x_i(t)$ of the single neurons are governed
by \footnote{~Note that we did not include external input signals in this
formula, as is done in \citep{FORCE} and \citep{RM}.}
\begin{equation}
    \dot{x}_i(t) = -x_i(t) + \sum_{i=1}^N \omega\Ix{rec}_{ij}\, r_j(t) + 
                   \sum_{i=1}^R \omega\Ix{fb}_{ij}\, z_j(t)~,
\end{equation}
where we introduce the internal recurrent synapse weights
$\omega\Ix{rec}_{ij}$ and the feedback synapse weights
$\omega\Ix{fb}_{ij}$. 
We solved this system of differential equations by
Euler integration with a time steps $dt = 0.01$, which was
also used in \citep{FORCE}. 
%(How we constructed $\omega\Ix{rec}$ and
%$\omega\Ix{fb}$ will be described in \emph{Parametric Random Networks})

\paragraph{Learning} Learning took place by stepwisely adapting the weights
$\omega\Ix{read}$ so that the dynamics $t\mapsto r_i(t)$ of the readout
neurons should eventually match a given periodic target pattern $t\mapsto
r'_i(t)$, henceforth called a \emph{task}, as accurately as possible. To
update $\omega\Ix{read}$ we chose variant A of the FORCE algorithm
\citep{FORCE}, which rapidly modifies the feedback loop to keep the errors
$|r_i(t) - r'_i(t)|$ small from the beginning on.

While the authors of \citep{FORCE} update the weights $\lambda\Ix{read}$ in
fixed intervals, we used an adaptive mechanism to
determine how often learning should occur. 
%This was initially introduced
%for performance reasons, as the single learning steps proved to be the main
%computational bottleneck, but we also found that the long time network
%dynamics tended to be slightly more accurate for adaptive learning steps.

\paragraph{Network Genotypes} Since we are not
interested in single network properties, but rather want to find out which
structural elements of networks are important for learning success, we
introduced  \emph{network genotypes}.
Such a genotype $G$ consists of a set $\Theta$ of parameters that determine
which concrete networks are created when $G$ is expressed in a single
phenotype. Technically $G$ may be understood as a stochastic network
generator that returns a network $\mathcal{N}=G(\lambda)$ for every seed
$\lambda \in \mathbb{N}$.
The genotypes' parameters $\theta\in\Theta$ are either integer or floating
point values and can be classified as affecting either (1) the topology of
$\omega\Ix{rec}$ or (2) the weight values of $\omega\Ix{rec}$ and
$\omega\Ix{fb}$. Important examples for (1) are the networks size $N$, the
occupation probability $p$ (for random Erdös-Renyi networks), and the
neighborhood range $k$ (for ring topologies). We furthermore introduced ratio parameters $r$
that allowed us to interpolate between different topologies. Optimized
parameters of type (2) were the feedback strength $\feedback$ (with
$\omega\Ix{fb}\propto \feedback$) and the gain value $\gain$ (with
$\omega\Ix{rec} \propto \gain$).

\paragraph{Fitness} The next step was to define the \emph{fitness} of a given
genotype when confronted with a diverse range of tasks, e.g. waves of
different frequencies. To express this formally, we introduced
\emph{challenges}: For a given seed value $\lambda$, the challenge $C$
yields a task $r' = C(\lambda)$ according to some (parametrized)
distribution. If we now make use of a success function $S(\mathcal{N},
r')\in\{0, 1\}$ that decides whether the NN $\mathcal{N}$
has successfully reproduced the desired output $r'$, the fitness $F_C(G)$
of a genotype $G$ for a challenge $C$ may reasonably be defined as the
expectation value of $\lambda \mapsto S(G(\lambda), C(\lambda))$. For fixed
challenges, the main task of optimizing the genotype is now reduced to
finding parameters $\Theta$ that maximize $F_C$.

@@

\paragraph{Choice of Parameters and Algorithmic Quantities}
The mutation rate of the parameters $\Theta$ of the genotypes\dots
